GeckoGraph: A Visual Language for Polymorphic Types

### 2025-10-21 main.tex
New changes

%% Abstract
\begin{abstract}
Polymorphic types are an important feature in most strongly typed programming languages. They allow functions to be written in a way that can be used with different data types, while still enforcing the relationship and constraints between the values. However, programmers often find polymorphic types difficult to use and understand and tend to reason using concrete types. We propose GeckoGraph, a graphical notation for types. GeckoGraph aims to accompany traditional text-based type notation and to make reading, understanding, and comparing types easier. We conducted a large-scale human study using GeckoGraph compared to text-based type notation. To our knowledge, this is the largest controlled user study on functional programming ever conducted. Overall, no significant difference were found  in time and success rate between the programmers use GeckoGraph and the ones who do not. However, we find that in the most difficult task of the user study, GeckoGraph improve the success rate for novice users (14.5\%) and less experienced users (11.2\%), suggesting the current iteration of GeckoGraph may be more useful as a teaching and learning tool. 
\end{abstract}

@@

While promoting robustness and code reusability, polymorphic types present challenges in learning and comprehension, particularly for novice users \cite{Jun2000-ec, Jun2000-yu}; as it is often argued that the expressiveness power of type systems often comes at the cost of their usability~\cite{Hage2020-hg}. Research indicates that humans tend to focus on concrete types and rely on polymorphic type checking only as a last resort. Moreover, polymorphic types often complicate type error usability. During type checking, compilers frequently generate new polymorphic type variables temporarily. These variables are typically discarded after type checking unless a type error occurs, in which case they are referenced in error messages. Consequently, programmers often face the task of resolving type errors involving type variables they did not explicitly author (Fig. \ref{fig:example-foldable}). Unfortunately, not many research focuses on the \textit{usability} of polymorphic types.

@@

 We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern Human-Computer Interaction (HCI) techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. We designed a gamified online study  \textit{ZeroToHero}\footnote{Available at https://zerotohero.fly.dev}, evaluating 714 participants' performance to solve a series of programming puzzles using GeckoGraph versus text-based type annotation, which is, to our knowledge, the largest controlled user study of a functional programming tool or concept.  Overall, no significant difference were found  in time and success rate between the programmers use GeckoGraph and the ones who do not. However, we find that in the hardest task of the user study, GeckoGraph improve the success rate for novice users (14.5\%) and less experienced users (11.2\%), suggesting the current iteration of GeckoGraph may be more useful as a teaching and learning tool.

@@

Static type systems are widely used in large and complex software projects because they help detect errors early and make refactoring safer. In functional programming languages, type systems are often more expressive and provide stronger safety guarantees. They enable programmers to model program behavior rigorously and enforce correctness at compile time. Common characteristics of type systems in functional programming include structural typing, algebraic data types, parametric polymorphism, and type inference.

@@

Unlike the nominal type systems found predominately in languages such as Java and C \cite{AbdelGawad2013-hh}, many functional languages heavily rely on structural typing. In nominal systems, two types are considered distinct even if they have identical structure but different names. In contrast, structural type systems consider types equivalent as long as their structures match, regardless of their names. This approach promotes flexibility, as types can be defined inline without introducing new names, and values are automatically valid for a type when they conform to the structure.  However, in functional languages such as Haskell, and OCaml, the lack of nominality can also make complex types harder to read and reason about. For instance, when encounter type errors, programmers can face two unmatching types with complex structure, understanding where exactly the root error is becomes a tediotasksks of string comparison.

@@

Algebraic data types (ADTs) \cite{Milner1978-qz} allow developers to construct complex types from simpler ones using sum and product operations. A sum types, often called tagged union, $A + B$ represent a type that can be a value of A or a value of B. Conversely, a product type $A \times B$ represent a tuple of A and B. This precise and composable type definitions for common data structures such as lists, tuples, and result types.
For example, in Haskell, types like \texttt{Either a b} and \texttt{Maybe a} provide basic building blocks for error handling, and the type \texttt{data List a = [] | a : List a } provides a succinct recursive definitions for lists. This illustrates both the expressive power and the potential complexity of ADTs. In practice, simple type definitions can turn into an \textit{essay of types} \cite{Ferdowsi2023-cv} as program grows, causing difficulties to interpret and maintain.

@@

Parametric polymorphism (often called generics) allows functions and data structures to operate uniformly across different types. For instance in Haskell, the\texttt{head :: [a] -> a}, taking a list as input and return the first element, is a polymorphic functions that works on lists of any type, integers, strings, any type is valid.  It provides a strong and principled form of abstraction, enabling code reuse without sacrificing type safety.

However, Parametric polymorphism also increases cognitive load: programmers must mentally track type variables and reason about how concrete types are instantiated during function application. Overuse can make code difficult to follow, especially for readers unfamiliar with the underlying type relationships.

@@

Type inference, or implicit typing, reduces the need for explicit type annotations by allowing the compiler to deduce types automatically.
Most modern languages employ some level of inference. For instance, Java’s var keyword (introduced in version 10) infers variable types at compile time based on their assigned values \cite{Java_Developers2023-an}, saving programmers from writing redundenies like \textit {Car car = new Car()}.

Functional languages such as Haskell and ML go further, using global type inference \cite{Damas1982-zw} to check entire programs without requiring explicit annotations. While this improves conciseness, it can reduce code readability and produce complex error messages. Many developers therefore choose to write explicit type signatures for clarity, even when inference is available.

To summarize, the type systems of functional programming languages provide rigorous tools for specifying and enforcing program behavior. By encoding system constraints directly in types, they prevent many classes of runtime errors. However, this rigor introduces trade-offs: a steeper learning curve and less flexibility for exploratory or prototype development. In the following sections, we demonstrate how GeckoGraph supports these advanced type-system features by visualizing type level information that is often implicit or difficult for humans to interpret.

@@

Motivated by the design goals, we designed GeckoGraph with the following feature features and construction rules (Fig. \ref{fig:design}):

@@

\paragraph{Complex types}
GeckoGraph displays an algebraic data type as a larger cell, where the type constructor half encloses its arguments: \includegraphics[height=1em]{figures/DataType.png}. The arguments are aligned in the bottom right of the cell. Two distinct visual dimensions are used to provide additional visual clarity (design goal \ref{goal2}). Data types containing more arguments (e.g., \texttt{ a b c} or \texttt{(a b) c}) will expand horizontally \includegraphics[height=1em]{figures/DataTypeWide.png}. Data types that are nested (e.g. \texttt{ a (b c)}) will grow taller \includegraphics[height=1.2em]{figures/DataTypeNested.png}. This distinction accommodates our design goal \ref{goal3}. Note that the height of GeckoGraph grows only upwards, but not downwards. Not only does this allow GeckoGraph to be more efficient in its space usage, but it also allows the legend text to be consistently aligned at the bottom and can be read as though it is a text-based type notation (design goal \ref{goal1}).

\paragraph{Function Types}
Functions are the fundamental building blocks of functional programming languages, and function types are ubiquitous and the most important in type-level programming. In Haskell, \texttt{(->)} is defined as an infix type operator with the right associativity to provide succinct type annotation. GeckoGraph preserves this syntax feature to make the notation more intuitive (design goal \ref{goal1}): the 2 arguments of a function type in GeckoGraph are placed on both sides of the cell \includegraphics[height=1em]{figures/Function.png}. A special function indicator (\texttt{$>>>$}) is displayed at the top of the cell.

Curried functions (e.g., \texttt{ a -> b -> c}) are displayed as two cells of functions merged together \includegraphics[height=1em]{figures/Curry.png}. It appears as if the second function covers the right half of the first function, indicating that the second function is the return type of the first. Regular high-order functions (e.g., \texttt{(a -> b) -> c}) follow the rules of functions and nested data types \includegraphics[height=1.2em]{figures/HOF.png}. The placement of function indicators aims to make it easy to find desired functions in the documentation site based on function arity and high-order functions (design goal \ref{goal3}). It is easy to tell high-order functions from the vertical position of its function indicator. Similarly, it is easy to count the arity of the outer function by counting the number of function indicators at the highest layer. For example, in Fig. \ref{fig:indicator}, the higest layer function indicator (the ones colored in the lightest shade of gray) suggest a ternary function. The two arguments types and the return type are all functions.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/Indicator}
  \caption{
        \label{fig:indicator}
        An example of visual hierarchy of the function indicator. The function indicator can be used to easily identify the arity of a function type by counting the connecting function indicators. For higher-order functions where functions are arguments or return values of other functions, it is very easy to see the ``level" of functions. 
  }
\end{figure}

@@

The GeckoGraph type class's design promotes the type class placements rather than the type class names. Programmers can easily see where and how many type classes are required, but they may need an extra step (global legends of the color mapping or pop-up window) to identify the name of the type class. We believe that this is a well-justified trade-off. The strength of this approach lies in situations when type class context is complex. When reading a type \texttt{(A a, A c, B a, B b, C b) => a -> b -> c}, programmers may need to switch back and forth to remember which type classes are needed on which variable. In GeckoGraph, type classes are explicitly displayed under the type variable cell,  minimizing programmers' effort to associate each type variable with all its type classes (design goal \ref{goal2}).

@@

\textbf{The colors of GeckoGraph} help programmers to see the permutation of type variables in the input and output of a function. A recent review \cite{Zeng2023-jz} of 59 graphical perception articles showed that combining solid color hue in a filled shape provides stronger visual perception for nominal data such as type identifiers. GeckoGraph uses qualitative color scheme from ColorBrewer \cite{Harrower2003-vu} to encode symbolic values of types. One example of GeckoGraph's use of color is the ``rotation" function  (Fig. \ref{fig:rotate}) in the user study (Section \ref{sec:evaluation}). With text-based type notation, programmers often rely on mnemonic devices such as alphabetic ordering or naming conventions. For example, the rotation function \texttt{f2 :: Zero a b c d -> Zero b c d a} appeared in the evaluation (Section \ref{sec:evaluation}), shuffles the type arguments one step to the left. The purpose of this function becomes less recognizable if changed to \texttt{f2 :: Zero e v m h -> Zero v m h e}.

\begin{figure}[]
  \includegraphics[width=0.6\linewidth]{figures/rotate}
  \caption{\label{fig:rotate} The `f2' function in the user study. It shuffles the type arguments one step to the left and the leftmost argument wraparound to be the rightmost.}
\end{figure}

\textbf{The horizontal axis of GeckoGraph} often are helpful in identifying differences in function arities. For example, in Fig. \ref{fig:add3}, the programmer intended to implement a function that sums 3 integers. In the implementation, the programmer missed a \texttt{(+)} function at the end; the resulting function type is largely different in length. It is also clear that the function needs to apply to one more binary function to satisfy the length requirement.

@@

\textbf{The vertical axis of GeckoGraph} often helps identify the number of levels in a nested complex type. This can often be very useful when inspecting mismatching type errors where data types are nested in different levels. Common examples include when programmers forget to apply the value to ``return" in a monadic block or to use \texttt{liftIO} to cast an \texttt{IO} effect. In Fig. \ref{fig:maybe}, the uses of \texttt{return} are excessive. It can be easily identified by examining the difference in the vertical layers of the two types. In text-based type notation, this is distinguished by different pairs of parentheses. However, parenthesis is an overloaded syntax in type notation. In Haskell, parentheses are used to enclose tuples \texttt{(a, b)}, specify the fixity \texttt{ (a -> b) -> c}, or have no effect \texttt{a -> (b -> c)}.

@@

\paragraph{Support Advanced Type Features}
The design of GeckoGraph enables the visualization of many advanced type-level features. \textbf{Kind visualization}: The kind system \cite{Jones1995-lm} is used in Haskell to indicate whether a type is a type constructor (e.g. [], Maybe) or a type constant (e.g. Bool, Char), because of this, kind is often described as the type of types. GeckoGraph infers the {\it kind} of type variables and consistently displays this information. For example, in Fig. \ref{fig:advanced} (A), the variable \texttt{a} is inferred to have the kind \texttt{* -> *} because of its use on the right-hand side. GeckoGraph respects this {\it kind} information and displays the same kind level on the first occurrence of \texttt{a}, but over an empty structure, indicated using a dotted outline. \textbf{Qualified constraints}: GeckoGraph's type class notation naturally extends to support qualified constraints. In the type \texttt{forall b. A (a b) => a b}, GeckoGraph shows the scope type class requirement on \texttt{a b} (Fig. \ref{fig:advanced} B).
\textbf{Multiple Parameter Type Class}:  GeckoGraph supports multiple parameter type classes by using multiple shapes with the same color hue to indicate the different parameters of the same type class. For example, for the type \texttt{A a b => a b},  GeckoGraph shows that the variables a and b both need an A class, but they are the different parameters of A (Fig. \ref{fig:advanced} C).

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/Advanced}
  \caption{\label{fig:advanced}
  Advanced features of GeckoGraph. (A) GeckoGraph supports Kind Visualization if the inferred kind is greater than \texttt{*}. (B) GeckoGraph supports qualified constraints by extending the extended area across multiple type variables. (C) GeckoGraph supports Multiple Parameter Type Classes, using different shapes of the same color to indicate that multiple variables must satisfy certain type classes collectively. (D) GeckoGraph supports the precise selection of its sub-structures. }
\end{figure}

@@

The experiment aims to study how programmers use and reason about polymorphic type when performing programming tasks. In particular, we studied how programmers scan and select potentially useful functions from a library and how they compare intended types and actual types when type errors are encountered.

@@

We illustrate the tasks of the user study using levels 2 and 3 as examples. For level 2 (Fig. \ref{fig:level-example}), the players need to implement the function \texttt{ zeroToHero :: Zero a -> Hero b}. The available functions are \texttt{runZero::Zero a -> a}, \texttt{makeHero:: a -> Hero a}. A generic function \texttt{(\$)} is provided to improve the ergonomics of composing functions, but all tasks can be completed without the use of generic functions. The possible solution and other details of the level can be found in Appendix \ref{levels}. To complete this level, players only need to use each of the provided function once. One possible solution to this task would be {\tt zeroToHero z = makeHero (runZero z)}. The user illustrated in Fig. \ref{fig:level-example} did not reach a correct solution, as the type signature and GeckoGraphs in {\it To Implement} and {\it Your Result} do not match.

\begin{figure}[!h]
  \includegraphics[width=\linewidth]{figures/Level3}
  \caption{\label{fig:level3} A screenshot showing the level 3 of the user study.  In this level, the players need to implement the function \texttt{ zeroToHero :: Zero a -> Hero (a a)}. Players are helped by the functions: {\tt f1 :: Zero a -> Hero a}, {\tt f2 :: Zero a -> (a, a)}, and {\tt f3 :: Hero a -> Hero (a, a)}.}
\end{figure}

@@

The experiment uses a between-subject design. Participants are assigned to one of two groups. The tasks assigned to the two groups are in the same order. Participants receive both treatments (with and without GeckoGraph) during their runs but in a different order.   Group one is assisted by GeckoGraph on even levels and text-only type annotation on odd levels. Group two participants are the same but with the order flipped. The number of participants in the two groups is counterbalanced.

@@

Participants can skip a level if they are stuck. We believe that it is normal for anyone to get stuck on a challenging task, and being stuck on one of the 10 tasks does not discount their feedback of the tool. The number of skips a participant can use is limited to four so that it is not possible to submit qualitative feedback without completing at least 6 tasks out of 10.

@@

Data collection from the study was stopped after the planned cut-off period of 14 days. After the cut-off date, the ZeroToHero game is open source and available for free evaluation\footnote{https://zerotohero.fly.dev} and repeating our experiment, but no further data was collected.

@@

The 10 levels are designed to increase difficulty gradually. From the results of the experiment, most of the tasks align with this trend. However, three tasks stand out in Fig. \ref{fig:level-time}.  Level 7 (mean = 334 seconds) is the hardest task in terms of time, followed by level 8 (mean = 228 seconds) and level 5 (mean = 224 seconds). To complete an average level, the beginner group uses an average of 100 seconds, the familiar group uses 90 seconds, the knowledgeable group uses 80 seconds, and the expert group uses 70 seconds. This roughly aligns with self-reported expertise. We show that the task time on each level follows normal distributions using a Shapiro-Wilk test \cite{Shaphiro1965-dx} (p-value  $ \leq 1.018 \times 10^-16$, for an alpha value of 0.05, p less than 0.05 is considered normal distribution).

Levels 5, 7, and 8 are the only three levels that include functions from standard Haskell library, baring the \texttt{(.)} and \texttt{(\$)} provided for convenience. Level 5 requires programmers to use the \texttt{fst} and \texttt{snd} functions to extract value from a tuple. Level 7 requires programmers to use the \texttt{(<*>)} function of the \texttt{Applicative} class, while level 8 requires the \texttt{fmap} function of the \texttt{Functor} class. The authors speculate that the more experienced participants are much more familiar with these functions, hence the strong contrast on these three levels.

However, when comparing the task time between the two treatments, we were unable to reject the null hypothesis. In a two-sample T-test, we could not find any significant difference between the two groups overall (p-value = 0.457), nor does there differ between the two groups in any of the four levels of experience (beginner: p-value = 0.845, familiar: p-value = 0.524, Knowledgeable: p-value = 0.712, expert p-value = 0.771).

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/LevelTime}
  \caption{\label{fig:level-time} Time spent on each level, with 95\% confidence interval. We show that the difficulty steadily increases across the user study, but levels 5, 7, and 8 are significantly harder than the authors intended. The overall task time of each group roughly matches experience level.}
\end{figure}

@@

% We start this section with a strength and weakness analysis of GeckoGraph. The disuccion is based on our results of user study. Then, In the end, we speculate the applications of GeckoGraph. The envisioned applications are grounded in the strengths and weaknesses of GeckoGraph. 
\section{Discussion} \label{sec:discussion}

>>>>

%% Abstract
\begin{abstract}
Polymorphic types are an important feature in most strongly typed programming languages. They allow functions to be written in a way that can be used with different data types, while still enforcing the relationship and constraints between the values. However, programmers often find polymorphic types difficult to use and understand, and tend to reason with a mindset of concrete types. We propose GeckoGraph, a graphical notation for types. GeckoGraph aims to accompany traditional text-based type notation and to make reading, understanding, and comparing types easier. We conducted a large-scale human study using GeckoGraph compared to text-based type notation. To our knowledge, this is the largest controlled user study on functional programming ever conducted. Our results show that no significant difference was found in time and success rate between the programmers who use GeckoGraph and those who do not. However, we also find that for the most challenging task of the user study, GeckoGraph improves the success rate by 14.5\% for beginners and 11.2\% for less experienced users, suggesting the current iteration of GeckoGraph may be more useful as a teaching and learning tool. 
\end{abstract}

@@

While promoting robustness and code reusability, polymorphic types present challenges in learning and comprehension, particularly for novice users \cite{Jun2000-ec, Jun2000-yu}; as it is often argued that the expressiveness power of type systems often comes at the cost of their usability~\cite{Hage2020-hg}. Research indicates that humans tend to focus on concrete types and rely on polymorphic type checking only as a last resort. Moreover, polymorphic types often complicate type error usability. During type checking, compilers frequently generate new polymorphic type variables temporarily. These variables are typically discarded after type checking unless a type error occurs, in which case they are referenced in error messages. Consequently, programmers often face the task of resolving type errors involving type variables they did not explicitly author (Fig. \ref{fig:example-foldable}). Unfortunately, not many studies focus on the \textit{usability} of polymorphic types.

@@

 We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern Human-Computer Interaction (HCI) techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. We designed a gamified user study \textit{ZeroToHero}\footnote{Available at https://zerotohero.fly.dev}, evaluating 714 participants' performance to solve a series of programming puzzles using GeckoGraph versus text-based type annotation, which is, to our knowledge, the largest controlled user study of a functional programming tool or concept.  Overall, no significant difference was found in either time or success rate between the programmers who use GeckoGraph and those who do not. However, we find that in the hardest task of the user study, GeckoGraph improves the success rate by 14.5\%  for novice users (less than 1 year of experience) and 11.2\% for less experienced users (1-3 years of experience), suggesting the current iteration of GeckoGraph may be useful as a teaching and learning tool.

@@

Functional programming languages have a rich history of embracing strong type systems and pioneering advanced type system features \cite{Milner1978-ob, MacQueen2020-qn, Hudak2007-kn}. In functional programming languages, type systems are often more expressive and provide stronger safety guarantees. They enable programmers to model program behavior rigorously and enforce correctness at compile time. Common characteristics of type systems in functional programming include structural typing, algebraic data types, parametric polymorphism, and type inference.

@@

Unlike the nominal type systems found predominantly in languages such as Java and C \cite{AbdelGawad2013-hh}, many functional languages heavily rely on structural typing. In nominal systems, two types are considered distinct even if they have identical structure but different names. In contrast, structural type systems consider types equivalent as long as their structures match, regardless of their names. This approach promotes flexibility, as types can be defined inline without introducing new names, and values are automatically valid for a type when they conform to the structure.  However, this lack of nominality can also make complex types harder to read and reason about. For instance, when encountering type errors, programmers can face two unmatching types with a complex structure, and understanding where exactly the root error is becomes a tedious task of string comparison.

@@

Algebraic data types (ADTs) \cite{Milner1978-qz} allow developers to construct complex types from simpler ones using sum and product operations. A {\bf sum type}, often called a tagged union, $A + B$, represents a type that can be a value of A or a value of B. Conversely, a {\bf product type} $A \times B$ represents a tuple type composed of two values of both {\tt A} and {\tt B}. This formalization allows precise and composable type construction for common data structures such as lists, tuples, and result types.
For example, in Haskell, types like \texttt{data Either a b = Left a | Right b} and \texttt{data Maybe a = Nothing | Just a} provide basic building blocks for error handling, and the type \texttt{data List a = [] | a : List a } provides a succinct recursive definition for lists. This illustrates the expressive power of ADTs. However, in practice, type definitions can grow to be complex structures, for instance \texttt {Maybe (Either [Maybe a] (Either a String))} as programs grow, causing difficulties to interpret and maintain.

@@

Parametric polymorphism (often called generics) allows functions and data structures to operate uniformly across different types. For instance, in Haskell, the function \texttt{head :: [a] -> a}, taking a list as input and returning the first element, is a polymorphic function that works on lists of any type; integers, strings, and functions are all valid candidates for the generic variable \texttt{a}.  It provides a strong and principled form of abstraction, enabling code reuse without sacrificing type safety.

However, parametric polymorphism also increases cognitive load: programmers must mentally track type variables and reason about how concrete types are instantiated during function application. This task becomes more challenging when functions involve multiple type variables. An example would be the foldM function: \texttt{foldM :: (Monad m) => (a -> b -> m a) -> a -> [b] -> m a}. The challenge of reading and understanding polymorphic types is more serious for novice users \cite{Jun2000-ec, Jun2000-yu}.

@@

Type inference, or implicit typing, reduces the need for explicit type annotations by allowing the compiler to deduce types automatically.
Most modern languages employ some level of inference. For instance, Java’s var keyword (introduced in version 10) infers variable types at compile time based on their assigned values \cite{Java_Developers2023-an}, saving programmers from writing redundant type declarations, such as \textit {Car car = new Car()}.

Functional languages such as Haskell and ML go further, using global type inference \cite{Damas1982-zw} to check entire programs without requiring explicit annotations. While this improves conciseness, it can reduce code readability and produce complex error messages. Many programmers, therefore, choose to write explicit type signatures for clarity, even when inference is available.

To summarize, the type systems of functional programming languages provide rigorous tools for specifying and enforcing program behavior. By encoding system constraints directly in types, they prevent many classes of runtime errors. However, this rigor introduces trade-offs: a steeper learning curve and less flexibility for exploratory or prototype development. In the following sections, we explore how GeckoGraph supports these type-system features by visualizing type-level information that is often implicit or difficult for humans to interpret.

@@

Motivated by the design goals, we designed GeckoGraph with the following features and construction rules (Fig. \ref{fig:design}):

@@

\paragraph{Algebraic Data Types}
GeckoGraph displays an algebraic data type as a larger cell, where the type constructor half encloses its arguments: \includegraphics[height=1em]{figures/DataType.png}. The arguments are aligned in the bottom right of the cell. Two distinct visual dimensions are used to provide additional visual clarity (design goal \ref{goal2}). Data types containing more arguments (e.g., \texttt{ a b c} or \texttt{(a b) c}) will expand horizontally \includegraphics[height=1em]{figures/DataTypeWide.png}. Data types that are nested (e.g. \texttt{ a (b c)}) will grow taller \includegraphics[height=1.2em]{figures/DataTypeNested.png}. This distinction accommodates our design goal \ref{goal3}. Note that the height of GeckoGraph grows only upwards, but not downwards. Not only does this allow GeckoGraph to be more efficient in its space usage, but it also allows the legend text to be consistently aligned at the bottom and can be read as though it is a text-based type notation (design goal \ref{goal1}).

\paragraph{Function Types}
Functions are the fundamental building blocks of functional programming languages, and function types are ubiquitous and the most important in type-level programming. In Haskell, \texttt{(->)} is defined as an infix type operator with right associativity to provide succinct type annotations. GeckoGraph preserves this syntax feature to make the notation more intuitive (design goal \ref{goal1}): the 2 arguments of a function type in GeckoGraph are placed on both sides of the cell \includegraphics[height=1em]{figures/Function.png}. A special function indicator (\texttt{$>>>$}) is displayed at the top of the cell.

Curried functions (e.g., \texttt{ a -> b -> c}) are displayed as two cells of functions merged together \includegraphics[height=1em]{figures/Curry.png}. It appears as if the second function covers the right half of the first function, indicating that the second function is the return type of the first. Regular high-order functions (e.g., \texttt{(a -> b) -> c}) follow the rules of functions and nested data types \includegraphics[height=1.2em]{figures/HOF.png}. The placement of function indicators aims to make it easy to find desired functions in the documentation site based on function arity and higher-order functions (design goal \ref{goal3}). It is easy to tell high-order functions from the vertical position of the function indicators. Similarly, it is easy to count the arity of a function by counting the number of function indicators at the outer layer. For example, in Fig. \ref{fig:indicator}, the outer layer function indicators (the ones colored in the lightest shade of gray) suggest a ternary function. The two argument types and the return type are all functions.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/Indicator}
  \caption{
        \label{fig:indicator}
        An example of the visual hierarchy of the function indicator. The function indicator can be used to easily identify the arity of a function type by counting the connecting function indicators. For higher-order functions where functions are arguments or return values of other functions, it is very easy to see the ``level" of functions. 
  }
\end{figure}

@@

The GeckoGraph type class's design promotes the type class placements rather than the type class names. Programmers can easily see where and how many type classes are required, but they may need an extra step (global legends of the color mapping or a pop-up window) to identify the name of the type class. We believe that this is a well-justified trade-off. The strength of this approach lies in situations when the type class context is complex. When reading a type \texttt{(A a, A c, B a, B b, C b) => a -> b -> c}, programmers may need to switch back and forth to remember which type classes are needed on which variable. In GeckoGraph, type classes are explicitly displayed under the type variable cell,  minimizing programmers' effort to associate each type variable with all its type classes (design goal \ref{goal2}).

@@

\textbf{The colors of GeckoGraph} help programmers to see the permutation of type variables in the input and output of a function. A recent review \cite{Zeng2023-jz} of 59 graphical perception articles showed that combining solid color hue in a filled shape provides stronger visual perception for nominal data such as type identifiers. GeckoGraph uses a qualitative color scheme from ColorBrewer \cite{Harrower2003-vu} to encode symbolic values of types. One example of GeckoGraph's use of color is the ``rotation" function  (Fig. \ref{fig:rotate}) in the user study (Section \ref{sec:evaluation}). With text-based type notation, programmers often rely on mnemonic devices such as alphabetic ordering or naming conventions. For example, the rotation function \texttt{f2 :: Zero a b c d -> Zero b c d a} appeared in the evaluation (Section \ref{sec:evaluation}), shuffles the type arguments one step to the left. The purpose of this function becomes less recognizable if changed to \texttt{f2 :: Zero e v m h -> Zero v m h e}.

\begin{figure}[]
  \includegraphics[width=0.6\linewidth]{figures/rotate}
  \caption{\label{fig:rotate} The `f2' function in the user study. It shuffles the type arguments one step to the left, and the leftmost argument wraps around to be the rightmost.}
\end{figure}

\textbf{The horizontal axis of GeckoGraph} is often helpful for identifying differences in function arities. For example, in Fig. \ref{fig:add3}, the programmer intended to implement a function that sums 3 integers. In the implementation, the programmer missed a \texttt{(+)} function at the end; the resulting function type is largely different in length. It is also clear that the function needs to apply to one more binary function to satisfy the length requirement.

@@

\textbf{The vertical axis of GeckoGraph} often helps identify the number of levels in a nested complex type. This can often be very useful when inspecting mismatched types where data types are nested at different levels. Common examples include when programmers forget to apply the value to ``return" in a monadic block or to use \texttt{liftIO} to cast an \texttt{IO} effect. In Fig. \ref{fig:maybe}, the uses of \texttt{return} are excessive. It can be easily identified by examining the difference in the vertical layers of the two types. In text-based type notation, this is distinguished by different pairs of parentheses. However, parenthesis is an overloaded syntax in type notation. In Haskell, parentheses are used to enclose tuples \texttt{(a, b)}, specify the fixity \texttt{ (a -> b) -> c}, or have no effect \texttt{a -> (b -> c)}.

@@

\paragraph{Support Advanced Type Features}
The design of GeckoGraph enables the visualization of many advanced type-level features. \textbf{Kind visualization}: The kind system \cite{Jones1995-lm} is used in Haskell to indicate whether a type is a type constructor (e.g., [], Maybe) or a type constant (e.g., Bool, Char). Because of this, kind is often described as the type of types. GeckoGraph infers the {\it kind} of type variables and consistently displays this information. For example, in Fig. \ref{fig:advanced} (A), the variable \texttt{a} is inferred to have the kind \texttt{* -> *} because of its use on the right-hand side. GeckoGraph respects this {\it kind} information and displays the same kind level on the first occurrence of \texttt{a}, but over an empty structure, indicated using a dotted outline. \textbf{Qualified constraints}: GeckoGraph's type class notation naturally extends to support qualified constraints. In the type \texttt{forall b. A (a b) => a b}, GeckoGraph shows the scope type class requirement on \texttt{a b} (Fig. \ref{fig:advanced} B).
\textbf{Multiple Parameter Type Class}:  GeckoGraph supports multiple parameter type classes by using multiple shapes with the same color hue to indicate the different parameters of the same type class. For example, for the type \texttt{A a b => a b},  GeckoGraph shows that the variables a and b both need an A class, but they are the different parameters of A (Fig. \ref{fig:advanced} C).

@@

\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/Advanced}
  \caption{\label{fig:advanced}
  Advanced features of GeckoGraph. (A) GeckoGraph supports Kind Visualization if the inferred kind is greater than \texttt{*}. (B) GeckoGraph supports qualified constraints by extending the extended area across multiple type variables. (C) GeckoGraph supports Multiple Parameter Type Classes, using different shapes of the same color to indicate that multiple variables must satisfy certain type classes collectively. (D) GeckoGraph supports the precise selection of its sub-structures. }
\end{figure}

@@

The experiment aims to study how programmers use and reason about polymorphic types when performing programming tasks. In particular, we studied how programmers scan and select potentially useful functions from a library and how they compare intended types and actual types when type errors are encountered.

@@

We illustrate the tasks of the user study using levels 2 and 3 as examples. For level 2 (Fig. \ref{fig:level-example}), the players need to implement the function \texttt{ zeroToHero :: Zero a -> Hero b}. The available functions are \texttt{runZero::Zero a -> a}, \texttt{makeHero:: a -> Hero a}. A generic function \texttt{(\$)} is provided to improve the ergonomics of composing functions, but all tasks can be completed without the use of generic functions. The possible solution and other details of the level can be found in Appendix \ref{levels}. To complete this level, players only need to use each of the provided functions once. One possible solution to this task would be {\tt zeroToHero z = makeHero (runZero z)}. The user illustrated in Fig. \ref{fig:level-example} did not reach a correct solution, as the type signature and GeckoGraphs in {\it To Implement} and {\it Your Result} do not match.

\begin{figure}[!h]
  \includegraphics[width=\linewidth]{figures/Level3}
  \caption{\label{fig:level3} A screenshot showing level 3 of the user study.  In this level, the players need to implement the function \texttt{ zeroToHero :: Zero a -> Hero (a a)}. Players are helped by the functions: {\tt f1 :: Zero a -> Hero a}, {\tt f2 :: Zero a -> (a, a)}, and {\tt f3 :: Hero a -> Hero (a, a)}.}
\end{figure}

@@

The experiment uses a between-subjects design. Participants are assigned to one of two groups. The tasks assigned to the two groups are in the same order. Participants receive both treatments (with and without GeckoGraph) during their runs, but in a different order.   Group one is assisted by GeckoGraph on even levels and text-only type annotation on odd levels. Group two participants are the same, but with the order flipped. The number of participants in the two groups is counterbalanced.

@@

Participants can skip a level if they are stuck. We believe that it is normal for anyone to get stuck on a challenging task, and being stuck on one of the 10 tasks does not discount their feedback on the tool. The number of skips a participant can use is limited to four, so that it is not possible to submit qualitative feedback without completing at least 6 tasks out of 10.

@@

Data collection from the study was stopped after the planned cut-off period of 14 days. After the cut-off date, the ZeroToHero game is open source and available for free evaluation\footnote{https://zerotohero.fly.dev}, and repeating our experiment, but no further data was collected.

@@

The 10 levels are designed to increase difficulty gradually. From the results of the experiment, most of the tasks align with this trend. However, three tasks stand out in Fig. \ref{fig:level-time}.  Level 7 (mean = 334 seconds) is the hardest task in terms of time, followed by level 8 (mean = 228 seconds) and level 5 (mean = 224 seconds). To complete an average level, the beginner group uses an average of 100 seconds, the familiar group uses 90 seconds, the knowledgeable group uses 80 seconds, and the expert group uses 70 seconds. This roughly aligns with self-reported expertise. We show that the task time on each level follows normal distributions using a Shapiro-Wilk test \cite{Shaphiro1965-dx} (p-value  $ \leq 1.018 \times 10^-16$, for an alpha value of 0.05, p less than 0.05 is considered a normal distribution).

Levels 5, 7, and 8 are the only three levels that include functions from the standard Haskell library other than the generic \texttt{(.)} and \texttt{(\$)}, which are provided for pure convenience. Level 5 requires programmers to use the \texttt{fst} and \texttt{snd} functions to extract a value from a tuple. Level 7 requires programmers to use the \texttt{(<*>)} function of the \texttt{Applicative} class, while level 8 requires the \texttt{fmap} function of the \texttt{Functor} class. The authors speculate that the more experienced participants are much more familiar with these functions, hence the strong contrast on these three levels.

However, when comparing the task time between the two treatments, we were unable to reject the null hypothesis. In a two-sample T-test, we could not find any significant difference between the two groups overall (p-value = 0.457), nor did there differ between the two groups in any of the four levels of experience (beginner: p-value = 0.845, familiar: p-value = 0.524, Knowledgeable: p-value = 0.712, expert: p-value = 0.771).

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/LevelTime}
  \caption{\label{fig:level-time} Time spent on each level, with 95\% confidence interval. We show that the difficulty steadily increases across the user study, but levels 5, 7, and 8 are significantly harder than the authors intended. The overall task time of each group roughly matches the experience level.}
\end{figure}

@@

\section{Discussion} \label{sec:discussion}
